Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation

### 2022-03-17 methods.tex
Changes to methods

The implicit solvent model can be described mathematically as a coupled system of partial differential equations, where the Poisson-Boltzmann governs in the solvent region ($\Omega_X$ in Figure XX), and the Poisson equation in the solute region ($\Omega_Y$ in Figure XX). These regions are interfaced by the molecular surface ($\Gamma$), where the potential ($\phi$) and electric displacement ($\epsilon\partial\phi/\partial\mathbf{n}$) are continuous. 
%
\begin{align}\label{eq:pbe}
\nabla^2\phi_1(\mathbf{x}) &= \frac{1}{\epsilon_1}\sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2(\mathbf{x})  &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1(\mathbf{x})  = \phi_2 (\mathbf{x}), &\quad \epsilon_1\frac{\partial\phi_1}{\partial\mathbf{n}}(\mathbf{x})  = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}}(\mathbf{x})  \quad \mathbf{x}\in \Gamma. 
\end{align}
%
where $\epsilon_1$ and $\epsilon_2$ are the dielectric constants in the solute and solvent, respectively, $\kappa$ is the inverse of the Debye length, related with the salt concentration, and $q_k$ are the values of the partial charges, located at $\mathbf{x}_k$.

@@

BEM reduces the dimension of the problem by using the boundary integral equation, hence its popularity as a numerical solution of  the Poisson-Boltzmann equation in molecular electrostatics. Unfortunately, such advantage comes also with the cost. A fundamental solution must be found before the BEM can be applied. There are many linear problems for which fundamental solutions are not known. It is a case if we consider the heterogenous permittivity inside the molecule
   \begin{align} \label{eq:pbe_vp}
\epsilon_1(\mathbf{x}) \nabla^2\phi_1(\mathbf{x}) &= \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2(\mathbf{x})  &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1(\mathbf{x})  = \phi_2(\mathbf{x}),  &\quad \epsilon_1(\mathbf{x})\frac{\partial\phi_1}{\partial\mathbf{n}}(\mathbf{x})  = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}}(\mathbf{x})  \quad \mathbf{x}\in \Gamma. 
\end{align}
For such case finite element methods (FEM) are more suitable.

@@

@@

@@

    The main difference between the standard and hybrid FEM-BEM formulations is addition of hybrid variable $\widetilde{\phi}$ that is a trace of $\phi_1$ and $\phi_2$ on the interface $\Gamma$ ($\widetilde{\phi} = \phi_1 = \phi_2$ on  $\Gamma$). Both FEM and BEM formulations are modified to include the hybrid variable by adding some terms that enforce the equality of it with the trace of $\phi_1$ and $\phi_2$.
    The previous standard formulation of interior problem~\eqref{eq:fem} can be rewritten as follows
\begin{align*}
 \int_{\Omega_m} \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} \nabla \phi_1 : \nabla v d\mathbf{x} -  \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2}  \partial_n \phi_1 v ds
  & \\
 -  \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} \partial_n v (\phi_1 - \widetilde{\phi}) ds
  + \tfrac{\tau_F}{h} \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} (\phi_1 - \widetilde{\phi}) v ds & = \int_{\Omega_1}  \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k)   v ~d\mathbf{x}.
\end{align*}
With every iteration we are solving above problem for given $\widetilde{\phi}$ .
%We are using FEniCS finite element library, hence the above formulation is more sumilar to the implementation.

In the case of exterior problem of~\eqref{eq:pbe_vp} we use Yukawa kernel and 
%matrix form, for clarity and similarity with BEMpp implementation. In this case we consider linearised formulation of the Poisson-Boltzman equation that means our operators are built as for modified Helmholtz equation with wavenumber $\kappa$.
For given $\widetilde{\phi}$, with each iteration we are solving the following system
\begin{align*}
V_{Y}^{\Gamma} \lambda  - \left( \tfrac{1}{2}Id + K_{Y}^{\Gamma}  \right) u &= - \widetilde{u},  \\
 \left( \tfrac{1}{2} Id +  K_{Y}^{',\Gamma}\right)  \lambda + \left(W_{Y}^{\Gamma} + \tau_B Id\right)  u  &= \tau_B  \widetilde{u},
\end{align*}
where $K'$ and $W$ are the adjoint double-layer potentials and hypersingular operator for the Yukawa kernel.

Both, the interior and exterior problems have unique solutions for given $\widetilde{u}$. In order to find $\widetilde{u}$ we solve the following problem on $\Gamma$
$$
\tau_B \left( u_s - \widetilde{u}\right) + \tfrac{\tau_F}{h} \left( u_m - \widetilde{u}\right) - \left(\frac{\epsilon_m}{\epsilon_s} \partial_n u -\lambda\right)  = 0.
$$
\end{itemize}

>>>>

The implicit solvent model can be described mathematically as a coupled system of partial differential equations, where the Poisson-Boltzmann governs in the solvent region ($\Omega_1$ in Figure XX), and the Poisson equation in the solute region ($\Omega_2$ in Figure XX). These regions are interfaced by the molecular surface ($\Gamma$), where the potential ($\phi$) and electric displacement ($\epsilon\partial\phi/\partial\mathbf{n}$) are continuous. 
%
\begin{align}\label{eq:pbe}
\nabla^2\phi_1(\mathbf{x}) &= \frac{1}{\epsilon_1}\sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2(\mathbf{x})  &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1(\mathbf{x})  = \phi_2 (\mathbf{x}), &\quad \epsilon_1\frac{\partial\phi_1}{\partial\mathbf{n}}(\mathbf{x})  = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}}(\mathbf{x})  \quad \mathbf{x}\in \Gamma. 
\end{align}
%
where $\epsilon_1$ and $\epsilon_2$ are the dielectric constants in the solute and solvent, respectively, $\kappa$ is the inverse of the Debye length, related with the salt concentration, and $q_k$ are the values of the partial charges, located at $\mathbf{x}_k$.

@@

BEM reduces the dimension of the problem by using the boundary integral equation, hence its popularity as a numerical solution of  the Poisson-Boltzmann equation in molecular electrostatics. Unfortunately, such advantage comes with the cost. A fundamental solution must be found before the BEM can be applied. There are many linear problems for which fundamental solutions are not known. For Poisson-Boltzmann equation it is a case if we consider the heterogenous permittivity inside the molecule
   \begin{align} \label{eq:pbe_vp}
\epsilon_1(\mathbf{x}) \nabla^2\phi_1(\mathbf{x}) &= \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2(\mathbf{x})  &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1(\mathbf{x})  = \phi_2(\mathbf{x}),  &\quad \epsilon_1(\mathbf{x})\frac{\partial\phi_1}{\partial\mathbf{n}}(\mathbf{x})  = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}}(\mathbf{x})  \quad \mathbf{x}\in \Gamma. 
\end{align}
For such case finite element methods (FEM) are more suitable.

@@

   As mentioned before we will use finite element discretisation for internal problem on domain $\Omega_1$ and boundary elements for external domain $\Omega_2$. The formulation below can be applied for homogeneous and heterogeneous $\epsilon_1$, hence for simplicity we do not distinguish between these cases in the formulation.

@@

Using the above definition and assuming that $\lambda_{1,\Gamma}  = \frac{\partial}{\partial \mathbf{n}} \phi_{1,\Gamma} $ the coupling problem can be written as follows
\begin{center}
  \textit{Find $ \phi_1 \in H^1(\Omega_1)$ and $\lambda_{1,\Gamma} \in H^{-\frac{1}{2}}(\Gamma)$ such that for all $v \in H^1(\Omega_1)$ and $\zeta \in H^{-\frac{1}{2}}(\Gamma)$}
\begin{equation} 
\label{eq:standard_fem_bem}
 \left\{
 \begin{array}{rcl}
 \int_{\Omega_1} \epsilon_1 \nabla \phi_1 : \nabla v ~d\mathbf{x}  - \oint_\Gamma \epsilon_1\partial_n \phi_1 v ds &=&   \int_{\Omega_1}  \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k)  v ~d\mathbf{x} \\[3mm] 
  \oint_\Gamma \left(\tfrac{1}{2} - K_{Y}^{\Gamma}\right) \phi_{1,\Gamma} \zeta ds + \frac{\epsilon_1}{\epsilon_2} \oint_\Gamma V_{Y}^{\Gamma} \lambda_{1,\Gamma} \zeta ds &=&0.
  \end{array}
  \right.
\end{equation}
\end{center}

%\begin{align}\label{eq:standard_fem_bem}
% \int_{\Omega_1} \epsilon_1 \nabla \phi_1 : \nabla v ~d\mathbf{x}  - \oint_\Gamma \epsilon_1\partial_n \phi_1 v ds &=   \int_{\Omega_1}  \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k)  v ~d\mathbf{x} \\
% \oint_\Gamma \frac{\phi_{1,\Gamma}}{2} - K_{Y}^{\Gamma}(\phi_{1,\Gamma}) + \frac{\epsilon_1}{\epsilon_2}V_{Y}^{\Gamma} \left( \lambda_{1,\Gamma} \right) & = 0
%\end{align}

@@

  In the previous section, we focused on the standard formulation. Now, we propose the weak penalty formulation of a couple problem. The discretisation is made by using finite element methods in the domain $\Omega_1$ and boundary integral methods in the domain $\Omega_2$.

    The main difference between the standard and hybrid FEM-BEM formulations is addition of hybrid variable $\widetilde{\phi}$ that is a trace of $\phi_1$ and $\phi_2$ on the interface $\Gamma$ ($\widetilde{\phi} = \phi_1 = \phi_2$ on  $\Gamma$). Both FEM and BEM subproblems are modified to include the hybrid variable by adding some terms that enforce the equality of it with the trace of $\phi_1$ and $\phi_2$.

    The standard formulation of interior problem~\eqref{eq:fem} can be rewritten as follows
\begin{align*}
 \int_{\Omega_m} \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} \nabla \phi_1 : \nabla v d\mathbf{x} -  \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2}  \partial_n \phi_1 v ds
  & \\
 -  \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} \partial_n v (\phi_1 - \widetilde{\phi}) ds
  + \tfrac{\tau_F}{h} \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} (\phi_1 - \widetilde{\phi}) v ds & = \int_{\Omega_1}  \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k)   v ~d\mathbf{x}.
\end{align*}
With every iteration we are solving above problem for given $\widetilde{\phi}$ and hence obtaining the solution $ \phi_1$.
%We are using FEniCS finite element library, hence the above formulation is more sumilar to the implementation.

In the case of exterior problem of~\eqref{eq:pbe_vp} we use Yukawa kernel and 
%matrix form, for clarity and similarity with BEMpp implementation. In this case we consider linearised formulation of the Poisson-Boltzman equation that means our operators are built as for modified Helmholtz equation with wavenumber $\kappa$.
for given $\widetilde{\phi}$, with each iteration we are solving the following system
\begin{align*}
V_{Y}^{\Gamma} \lambda_{2,\Gamma}  - \left( \tfrac{1}{2}Id + K_{Y}^{\Gamma}  \right) \phi_2 &= - \widetilde{\phi},  \\
 \left( \tfrac{1}{2} Id +  K_{Y}^{',\Gamma}\right)  \lambda_{2,\Gamma} + \left(W_{Y}^{\Gamma} + \tau_B Id\right)  \phi_2  &= \tau_B \widetilde{\phi},
\end{align*}
where $K'$ and $W$ are the adjoint double-layer potentials and hypersingular operator for the Yukawa kernel.

Both, the interior and exterior problems have unique solutions for given $\widetilde{\phi}$. In order to find $\widetilde{\phi}$ we solve the following problem on $\Gamma$
\begin{equation*}
\tfrac{\tau_F}{h} \left( \phi_1 - \widetilde{\phi}\right) + \tau_B \left( \phi_2 - \widetilde{\phi}\right)  - \left(\frac{\epsilon_1}{\epsilon_2} \frac{\partial}{\partial \mathbf{n}} \phi_1 -\lambda_{2,\Gamma} \right)  = 0.
\end{equation*}
\end{itemize}
